---
title: "Energy Transitions R Tutorial"
output: 
  html_document:
    keep_md: true
---

Using R, create a plot of primary energy consumption (Quadrillion) versus time from 1635 -2017.

## Read in CSV file

In base R, you can read in data as "dataframes" very easily. However, I would like to get you all in the habit of using "data tables" from the "data.table" library. It's great for later on (should you continue to use R for data analysis) when you will likely encounter larger datasets. Data tables are much more time-efficient to process on than dataframes.

First, let's install the package:
```{r}
# install.packages("data.table")
```
Then, we can load the data.table library:
```{r} 
library(data.table)
```

Let's now read in the CSV file. If you open the file itself, you can see that the first two rows are not actually data we need. We can specify our code to skip the first two rows.
```{r}
df_data = fread('data/Primary Energy Consumption_from 1635.csv', skip = 2)
```

Click on the *df_data* object in your work environment to view the entire data table, or run the following code to see the "head" (first 6 rows) of the data table:
```{r}
head(df_data)
```

## Rename year column

The first column, which includes the years, was not assigned a column name since the CSV file does not have a column name in that cell. So, we can manually change the first column as such:
```{r}
colnames(df_data)[1] = 'year'
head(df_data)
```

## Fill NA values

There are a lot of records that are null (NA). Let's set these as equal to zero:
```{r}
df_data[is.na(df_data)] = 0 
```

## Create new column that's a sum of other columns

Let's create a new column titled 'Other Renewables' that is a sum of Solar and Wind fuel consumption.
```{r}
df_data[, 'Other Renewables' := Solar + Wind]
tail(df_data)
```

## Transform data

When plotting a time series of categorical data, it's often easiest to have your data in a "long" format instead of a "wide" format (like above). So, this next step will transform your data so that all the values for the different fuel types will be put into a single column. Another column labeled "Fuel" will specify the type of fuel.

```{r}
df_long = melt(df_data, id.vars = "year", measure.vars = c(2:12),
               variable.name = "fuel", value.name = "value")
head(df_long)
```

## Remove certain variables 

Let's remove rows where the fuel are the totals ("Total Fossil", "Total Renewable Energy") as well as "Solar" and "Wind" since they're already included in the "Other Renewables" variable. 
The *%in%* selects all rows where the "Fuel" column included in the vector *c("Total Fossil", "Total Renewable Energy", "Solar", "Wind")*. The exclamation point (!) is equivalent to a *not* in the code. Thus, the code is actually selecting all rows where the fuel does NOT equal the listed fuels: 
```{r}
df_fuels = df_long[! fuel %in% c("Total Fossil", "Total Renewable Energy", "Solar", "Wind")]
```

Alternatively, you can specify the rows with the fuel types you **do** want to keep:
```{r}
df_fuels = df_long[ fuel %in% c("Coal", "Natural Gas", "Petroleum", "Nuclear", "Hydropower", "Wood/biomass", "Other Renewables")]
```
You can see that the resulting data table are the same either way.

## Create plot of primary fuel consumption over time

### Load ggplot
*ggplot* is a very popular package in R used to create plots. It's incredible versatile and flexible, and we'll be using the package to create the plot.
First, let's install the package:
```{r}
# install.packages("ggplot2")
```

```{r}
library(ggplot2)
```
### Rough plotting
```{r}
plot_cons = ggplot(data = df_fuels, aes(x = year, y = value, color = fuel)) + geom_line()
plot_cons
```

### Change axes limits and range

Let's change the title, labels, and axes of the plot.
```{r}
plot_cons = plot_cons + 
  scale_x_continuous(limits = c(1635,2018), breaks = seq(1600,2000,50), expand = c(0,0)) +
  scale_y_continuous(limits = c(0,45), breaks = seq(0,45,5)) 
plot_cons
```

### Specify line colors
```{r}
plot_cons = plot_cons + scale_color_manual(values = c("Petroleum" = "orange",
                                              "Natural Gas" = "red",
                                              "Coal" = "black",
                                              "Nuclear" = "seagreen",
                                              "Hydropower" = "steelblue",
                                              "Wood/biomass" = "brown",
                                              "Other Renewables" = "skyblue"))
plot_cons
```

### Add title, subtitle, and caption
Be sure to change Prof. Sanders' name to your own and to change the date as well.
```{r}
plot_cons = plot_cons + labs(title = "U.S. Primary Energy Consumption by Source (1635 – 2018)", 
                     subtitle = "Quadrillion British Thermal Units",
                     caption = "Plot created by Kelly Sanders on August 21, 2019",
                     x = NULL,
                     y = NULL,
                     color = NULL) 
plot_cons
```

### Making the plot prettier

I'm not a huge fan of how the plot looks right now, so let's pretty it up a bit, shall we? First up, let's make it a more minimal plot and change the font to Arial Narrow:
```{r}
  plot_cons = plot_cons + theme_minimal(base_family = "Arial Narrow")
  plot_cons
```

Better but can still be improved. Let's change the font sizes and styles a bit.
```{r}
  plot_cons = plot_cons + theme(plot.title = element_text(size = 16, face = "bold"),
                        plot.subtitle = element_text(size = 12, face = "plain"),
                        plot.caption = element_text(size = 10, face = "italic"),
                        axis.title = element_text(size = 14, face = "bold"),
                        axis.text = element_text(size = 12, face = "plain"),
                        legend.text = element_text(size = 12, face = "plain"))
  plot_cons
```

Okay, now let's get rid of some of the grid lines so that there's only a y-axis line.
```{r}
  plot_cons = plot_cons + theme(panel.grid = element_line(color = "#cccccc", size = 0.2),
                        panel.grid.major = element_line(color = "#cccccc", size = 0.2),
                        panel.grid.major.x = element_blank(),
                        panel.grid.minor = element_blank(),
                        plot.margin = margin(10, 10, 10, 10),
                        panel.spacing = grid::unit(2, "lines")) 
  plot_cons
```

### Saving the line plot as an image file

To see where your current working directory is, use the *getwd()* function (usually, if you're running this notebook in RStudio, it should already be the location where the notebook is saved on your computer:
```{r}
getwd()
```

If that's not where you want your image files to be saved, you can use the *setwd()* function to change it. 

To save the plot (in the sub-folder named "figs"):
```{r}
  ggsave(plot_cons, 
         path = "figs",
         filename = "Figure_Fuel-Consumption-by-Source_KTSanders.png", 
         width = 8.2, 
         height = 4.4, 
         dpi = 600)
```
Be sure to change the filename to your own.


### Removing the legend completely

This next step isn't very necessary, but let's say you don't like having a legend there, and instead you want the fuel source labels to be directly next to the line plot. To do so, first load the following the libraries:
```{r}
library(directlabels)
library(grid)
```

```{r}
  plot_cons_cut = plot_cons + 
    geom_dl(aes(label = fuel), method = list(dl.trans(x = x + .3), "last.bumpup", cex = 0.8)) +
    guides(color = FALSE) +     
    theme(plot.margin = margin(10, 110, 10, 10))
```

The following chunk cuts the plot and adds labels on top
```{r}
  plot_cons_cut = ggplotGrob(plot_cons_cut)
  plot_cons_cut$layout$clip[plot_cons_cut$layout$name == "panel"] = "off"
  # grid.draw(plot_cons_cut)
```

We can save the new plot: 
```{r}
  ggsave(plot_cons_cut, 
         path = "figs",
         filename = "Figure_Fuel-Consumption-by-Source_KTSanders_V2.png", 
         width = 8.2, 
         height = 4.4, 
         dpi = 400)
```


## Data analysis

### General analysis

We can play around with the data to do some rough analyses.
First, let's see which fuel and year had the highest value of annual fuel consumption:
```{r}
df_fuels[, .SD[which.max(value)]]
```
So, the largest amount of fuel consumption (in units of quadrillion BTUs) was petroleum consumption in the year 2005. 

But what if we want to know the maximum amount (and the corresponding year) consumed *for each fuel type*? Luckily, we can easily group our function to find the max:
```{r}
df_fuels[, .SD[which.max(value)], by = 'fuel']
```
We can see that coal consumption peaked in 2005, whereas consumption of natural gas and non-hydro renewables peaked just last year in 2018.

Let's sum up fuel consumption every year, across all fuel types:

```{r}
agg_annual = df_fuels[, .(value = sum(value)), by = year] # sum fuel consumption by year
```

When was our primary fuel consumption the greatest? Just last year in 2018, it seems.
```{r}
agg_annual[, .SD[which.max(value)]]
```
In fact, over the past 10 years, our total primary fuel consumption has been increasing.
```{r}
tail(agg_annual, 10)
```

But we can see from the plots we created that coal consumption has been on the decline, while natural gas and non-hydro consumption have been increasing.

### 10-year changes

To create a calculation of rate of change every 10 years, let's create a sequence of years that we want to include. In this case, we're going to create a 10 year sequence from 1635 to 2015.
```{r}
# year_seq = seq(max(df_fuels[, year]), min(df_fuels[, year]), -10)
year_seq = seq(min(df_fuels[, year]), max(df_fuels[, year]), 10)
year_seq
```

Create a new data table to include only those years:
```{r}
agg_10yr = df_fuels[year %in% year_seq]
head(agg_10yr)
```

Let's make a new column telling us what the span of years are. Intuitively, we know the years are 10 years apart. 
```{r}
agg_10yr[, prev_year := year-10]
agg_10yr[, span := paste(prev_year, year, sep = " - ")]
head(agg_10yr)
```

Calculate the difference and percent difference between each row and the row before it, grouped by fuel type.
```{r}
agg_10yr[ , diff := value - shift(value), by = fuel]    
agg_10yr[ , perc_diff := (value - shift(value))/shift(value), by = fuel]    
head(agg_10yr)
```

Some of the columns are unncessary, so let's just clean up the data table a bit:
```{r}
agg_10yr = agg_10yr[, c("fuel", "span", "year", "value", "diff", "perc_diff")]
head(agg_10yr)
```

What was the greatest (maximum) change in absolute value of quadrillion BTUs consumed over a ten year span? 
```{r}
agg_10yr[!is.na(diff), .SD[which.max(diff)]]
```
Between 1965 and 1975, petroleum consumption increased by almost 9.5 quadrillion BTUs (amounting to a 41% change).

Okay, but what if we want to know what was the greatest absolute change in fuel consumption *for each fuel type*? 
```{r}
agg_10yr[!is.na(diff), .SD[which.max(diff)], by = 'fuel']
```
We can see that coal experienced the greatest absolute growth in the years 1935 - 1945. Non-hydro renewables experienced the greatest growth in consumption from 2005 to 2015.

On the other hand, we can also when the most drastic decreases in fuel consumption was by singling out rows with a negative difference, then finding the minimum.
```{r}
agg_10yr[!is.na(diff) & diff < 0, .SD[which.min(diff)], by = fuel]
```
Coal experienced the biggest drop in absolute fuel consumption within the past decade, with an over 7 quadrillion decrease (~32% decrease). 

We can analyze percentage differences (difference relative to the original year) as well. For starters, let's look at when each fuel type experienced the greatest percent increase:
```{r}
agg_10yr[is.finite(perc_diff), .SD[which.max(perc_diff)], by = 'fuel']
```
Now, our results have changed a bit. Because we're using a percent difference, the data is skewed towards earlier years where there wasn't much fuel consumed (within each fuel type) anyways. For example, non-hydro renewables experienced a 588% increase in fuel consumption between 1985 - 1995 That can seem like a lot, but when you look at how in 1995, only 0.100863 quadrillion BTUs of non-hydro renewable was consumed, it's still a very small amount. 

What if we want to see the most drastic percent decrease?
```{r}
agg_10yr[is.finite(perc_diff) & perc_diff < 0, .SD[which.min(perc_diff)], by = fuel]
```

## Plotting percent changes

### Line plot

```{r}
line_diff = ggplot(data = agg_10yr, aes(x = year, y = diff, color = fuel, group = fuel)) + 
  geom_line() +
  labs(title = "10-Year Change in U.S. Primary Energy Consumption by Source (1635 – 2015)", 
       subtitle = "Quadrillion British Thermal Units",
       caption = "Plot created by Kelly Sanders on August 21, 2019",
       x = NULL,
       y = NULL) +
  scale_color_manual(values = c("Petroleum" = "orange",
                               "Natural Gas" = "red",
                               "Coal" = "black",
                               "Nuclear" = "seagreen",
                               "Hydropower" = "steelblue",
                               "Wood/biomass" = "brown",
                               "Other Renewables" = "skyblue")) +
  guides(color = FALSE) +
  theme_minimal(base_family = "Arial Narrow") +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12, face = "plain"),
        plot.caption = element_text(size = 10, face = "italic"),
        axis.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 12, face = "plain"),
        legend.text = element_text(size = 12, face = "plain"),
        panel.grid = element_line(color = "#cccccc", size = 0.2),
        panel.grid.major = element_line(color = "#cccccc", size = 0.2),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        plot.margin = margin(10, 10, 10, 10),
        panel.spacing = grid::unit(2, "lines"))
line_diff
```


```{r}
line_diff_facet = ggplot(data = agg_10yr, aes(x = year, y = diff, color = fuel, group = fuel)) + 
  geom_line() +
  facet_wrap(~fuel, ncol = 2) +
  labs(title = "10-Year Change in U.S. Primary Energy Consumption by Source (1635 – 2015)", 
       subtitle = "Quadrillion British Thermal Units",
       caption = "Plot created by Kelly Sanders on August 21, 2019",
       x = NULL,
       y = NULL) +
  scale_color_manual(values = c("Petroleum" = "orange",
                                "Natural Gas" = "red",
                                "Coal" = "black",
                                "Nuclear" = "seagreen",
                                "Hydropower" = "steelblue",
                                "Wood/biomass" = "brown",
                                "Other Renewables" = "skyblue")) +
  guides(color = FALSE) +
  theme_minimal(base_family = "Arial Narrow") +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12, face = "plain"),
        plot.caption = element_text(size = 10, face = "italic"),
        axis.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 12, face = "plain"),
        legend.text = element_text(size = 12, face = "plain"),
        panel.grid = element_line(color = "#cccccc", size = 0.2),
        panel.grid.major = element_line(color = "#cccccc", size = 0.2),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        plot.margin = margin(10, 10, 10, 10),
        panel.spacing = grid::unit(2, "lines"))
line_diff_facet
```

### Bar plots

```{r}
bar_diff_facet = ggplot(data = agg_10yr, aes(x = year, y = diff, fill = fuel, group = fuel)) + 
  geom_bar(stat = "identity") +
  facet_wrap(~fuel, ncol = 2) +
  labs(title = "10-Year Change in U.S. Primary Energy Consumption by Source (1635 – 2015)", 
       subtitle = "Quadrillion British Thermal Units",
       caption = "Plot created by Kelly Sanders on August 21, 2019",
       x = NULL,
       y = NULL,
       color = NULL) +
  scale_fill_manual(values = c("Petroleum" = "orange",
                                "Natural Gas" = "red",
                                "Coal" = "black",
                                "Nuclear" = "seagreen",
                                "Hydropower" = "steelblue",
                                "Wood/biomass" = "brown",
                                "Other Renewables" = "skyblue")) +
  guides(fill = FALSE) +
  theme_minimal(base_family = "Arial Narrow") +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12, face = "plain"),
        plot.caption = element_text(size = 10, face = "italic"),
        axis.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 12, face = "plain"),
        legend.text = element_text(size = 12, face = "plain"),
        panel.grid = element_line(color = "#cccccc", size = 0.2),
        panel.grid.major = element_line(color = "#cccccc", size = 0.2),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        plot.margin = margin(10, 10, 10, 10),
        panel.spacing = grid::unit(2, "lines"))
bar_diff_facet
```

